---
title: "cor_and_plot"
author: "Dorcas Washington"
date: "6/3/2020"
output: html_document
---

# Correlation Plots

In the Data Analysis With R workshop I waws asked to dig into correlation plots. I'm still not sure if you can visualization correlation matrices with just R base (e.g. without loading any other packages) but I do know of some easy (3-5 short lines of code) you can run to get correlation visualizations.

In the first step below we are running the setup chunk. 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Using the 'psych' package 

```{r}
## Load data
install.packages('psych')
mtcars <- mtcars

#subset data to variables of interest (I just chose 5 variables)
mydata <- mtcars[, c('mpg', 'cyl', 'disp', 'hp', 'carb')]
library(psych)
pairs.panels(mydata)

```

If you'd like to see the correlation between all the data the code below does that,

```{r}
#you can look at all the variables but it looks a crowed and messy in my opinion and might be hard to interpret even for exploratory analysis 
pairs.panels(mtcars)

```


## Using 'corrplot' and 'RColorBrewer' packages


```{r}
install.packages(c('corrplot', 'RColorBrewer'))
library(corrplot)
library(RColorBrewer)
corrplot.mixed(cor(mtcars), order="hclust", tl.col="black", diag = "l")

#changing the the lower diagonal text color to black so you can read it
corrplot.mixed(cor(mtcars), order="hclust", tl.col="black", diag = "l", lower.col = "black")

#notice how it's still hard to read with all the variables in the matrix


# using less variables it's easier to see so there's trade offs
mydata <- mtcars[, c('mpg', 'cyl', 'disp', 'hp', 'carb')]

corrplot.mixed(cor(mydata), order="hclust", tl.col="black", diag = "l", lower.col = "black")

```

